A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-19 paper.Rmd
cleaning ...

The impact of ambient temperature variations on infant mortality is studied for 
a northern Sweden coastal area, the Umeå and Skellefteå regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and climate variations in the same geographical area during the 
years 1880--1950. Climate and mortality in general is a research area that has 
generated great interest over the last years, see @tbgb10.

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with sex, social class, and legitimacy. Studies are performed 
separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
and in a joint analysis, the classification into endogenous and exogenous factors will be discussed.

One important reason for studying neonatal and postneonatal mortality separately is the 
findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
so we start by recollecting his results and formulate them in a modern statistical setting.

@@

Bourgeois-Pichat (*B-P*) suggests that mortality
during the last eleven months (28 days) of infancy follows a *uniform* distribution
given a specific transformation of time (age), the 
*log-cube transformation* $g$.

@@

Furthermore, it is assumed that the *endogeneous* component is non-zero only during the first
month of life.

@@

The result is shown in Figure \@ref(fig:ggplot2), together with the original
data.  The transform data is stretced  stretched to 365 for easier comparison.

@@

We create the neonatal data set by censor all lives at the age 28:

@@

and then fit a piecewise constant hazards model with cutpoints at $g(0, 2, 4, 7, 11, 16, 22, 28)$, see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", axes = FALSE)
axis(1)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

@@

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 3}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
par(las = 1)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(5, 39))
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

@@

```{r together, fig.cap = "Hazard function, infant mortality, transformed time."}
allcuts <- c(neocuts, g(365))
ginfdat <- infdat
ginfdat$exit <- g(ginfdat$exit)
fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = allcuts)
plot(fitall, fn = "haz", col = "blue", main = "", xlab = "g(Days)", axes = FALSE,
     ylim = c(0, 0.002))
pcuts <- g(c(0, 28, 183, 365))
axis(1, at = pcuts, labels = round(pcuts))
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

@@

```{r postanal, results='asis'}
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951))
levels(infdat$subreg) <- c("ume", "ske")
infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
infdat$season <- as.factor(infdat$season)
pneo <- age.window(infdat, c(28 / 365, 1))
##
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
pneo$emintemp <- round(pneo$emintemp)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
                    period + cold
                + emintemp + season + subreg, 
                data = pneo, cuts = g(c(28, 365)))
##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, shape = 1)
##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, cuts = g(c(28, 365)))
fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity +
                        period + cold + emintemp + season + subreg, 
                    data = tpneo, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality.", label = "tab:postanal")
```

>>>>

The impact of ambient temperature variations on infant mortality is studied for 
a northern Sweden coastal area, the Umeå and Skellefteå regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and temperature variations in a larger geographical area containing 
the present one during the 
years 1880--1950. Climate and mortality in general is a research area that has 
generated great interest over the last years, see @tbgb10.

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with sex, social class, and legitimacy. Studies are performed 
separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
and the classification into endogenous and exogenous factors will be discussed.

One important reason for studying neonatal and postneonatal mortality separately is the 
empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
so we start by recollecting his results and formulate them in a modern statistical setting.

@@

Bourgeois-Pichat (*B-P*) suggests that mortality
during the last eleven months of infancy follows a *uniform* distribution
given a specific transformation of time (age), the 
*log-cube transformation* $g$, see equation \@ref(eq:g),

@@

Furthermore, it is assumed that the *endogeneous* component is non-zero only during the first
28 days of life.

@@

The result is shown in Figure \@ref(fig:ggplot2), together with the original
data.  The transform data is stretched  stretched to 365 for easier comparison.

@@

We create the neonatal data set by censor all lives at the age 28 days:

@@

```{r defcuts}
neocuts <- c(0, 2, 5, 8, 12, 17, 22, 25, 28)
pcuts <- c(neocuts, 365)
gneocuts <- g(neocuts)
gpcuts <- g(pcuts)
```

and then fit a piecewise constant hazards model with cutpoints at `r gneocuts`, see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = gneocuts)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE, lwd = 1.5)
axis(1, at = gneocuts, labels = neocuts)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

@@

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 3}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
par(las = 1)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", lwd = 1.5,
     xlim = g(c(5, 28)), axes = FALSE)
axis(1, at = gneocuts, labels = neocuts)
axis(2)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2, lwd = 1.5)
```

@@

```{r together, fig.cap = "Hazard function, infant mortality, transformed time.", fig.height=6}
ginfdat <- infdat
ginfdat$exit <- g(ginfdat$exit)
fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = gpcuts)
plot(fitall, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
     lwd = 1.5)

axis(1, at = gpcuts, labels = pcuts)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

@@

```{r postanal, results='asis'}
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
levels(infdat$subreg) <- c("ume", "ske")
infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
infdat$season <- as.factor(infdat$season)
pneo <- age.window(infdat, c(28 / 365, 1))
##
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
pneo$emintemp <- round(pneo$emintemp)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
                    period + cold
                + emintemp + season + subreg, 
                data = pneo, cuts = g(c(28, 365)))
##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, shape = 1)
##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, cuts = g(c(28, 365)))
fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity +
                        period + cold + emintemp + season + subreg, 
                    data = tpneo, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality.", label = "tab:postanal")
```
